---
title: "nestedcv"
author: "Myles Lewis"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: true
fig_width: 6
vignette: >
  %\VignetteIndexEntry{nestedcv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(nestedcv)
```

# Introduction

Nested cross-validation (CV) for the glmnet and caret packages. With glmnet this
includes cross-validation of elastic net alpha parameter. A number of filter
functions (t-test, wilcoxon test, anova, Pearson/Spearman correlation, random
forest, ReliefF) for feature selection are provided and can be embedded within
the outer loop of the nested CV. Nested CV can be also be performed with the
caret package giving access to the large number of prediction methods available
in caret.

# Installation

Install from Github (requires API token).

```{r eval = FALSE}
devtools::install_github("myles-lewis/nestedcv", auth_token = "API token...")
library(nestedcv)
```

# Nested CV with glmnet

Using R4RA data to predict CDAI 50% response to rituximab.

```{r eval = FALSE}
# Data for this example can be downloaded from:
# DOI: 10.6084/m9.figshare.19336679

# set up data
load("/../R4RA_270821.RData")

index <- r4ra.meta$Outliers_Detected_On_PCA != "outlier" & r4ra.meta$Visit == 3 
          & !is.na(r4ra.meta$Visit)
metadata <- r4ra.meta[index, ]
dim(metadata)  # 133 individuals

medians <- Rfast::rowMedians(as.matrix(r4ra.vst))
data <- t(as.matrix(r4ra.vst))
# remove low expressed genes
data <- data[index, medians > 6]  
dim(data)  # 16254 genes

# Rituximab cohort only
yrtx <- metadata$CDAI.response.status.V7[metadata$Randomised.medication == "Rituximab"]
yrtx <- factor(yrtx)
data.rtx <- data[metadata$Randomised.medication == "Rituximab", ]

# no filter
res.rtx <- nestcv.glmnet(y = yrtx, x = data.rtx, min_1se = 0.5,
                         family = "binomial", cores = 8,
                         alphaSet = seq(0.7, 1, 0.05))
res.rtx
```

Use `summary()` to see full information from the nested model fitting. `coef()`
can be used to show the coefficients of the final fitted model.
Filters can be used by setting the `filterFUN` argument. Options for the filter 
function are passed as a list through `filter_options`.

```{r eval = FALSE}
# t-test filter
res.rtx <- nestcv.glmnet(y = yrtx, x = data.rtx, min_1se = 0, filterFUN = ttest_filter,
                         filter_options = list(nfilter = 300, p_cutoff = NULL),
                         family = "binomial", cores = 8,
                         alphaSet = seq(0.7, 1, 0.05))
summary(res.rtx)
```

Output from the nested CV with glmnet can be plotted to show how deviance is 
affected by alpha and lambda.

```{r eval = FALSE}
plot_alphas(res.rtx)
plot_lambdas(res.rtx)
```

The tuning of alpha for each outer fold can be plotted.

```{r eval = FALSE}
plot(res.rtx$outer_result[[1]]$cvafit)

# scatter plot
plot(res.rtx$outer_result[[1]]$cvafit, type = 'p')

# number of non-zero coefficients
plot(res.rtx$outer_result[[1]]$cvafit, xaxis = 'nvar')
```

ROC curves from left-out folds from both outer and inner CV can be plotted.

```{r eval = FALSE}
# Outer CV ROC
plot(res.rtx$roc)

# Inner CV ROC
rtx.inroc <- innercv_roc(res.rtx)
plot(rtx.inroc)
pROC::auc(rtx.inroc)
```

The overall expression level of each gene selected in the final model can be 
compared with a boxplot.

```{r eval = FALSE}
boxplot_model(res.rtx, data.rtx, ylab = "VST")
```

Other filters include wilcoxon (Mann-Whitney) test, Pearson or Spearman 
correlation for regression modelling, random forest and ReliefF filters.

```{r eval = FALSE}
# random forest filter
res.rtx <- nestcv.glmnet(y = yrtx, x = data.rtx, min_1se = 0.5, filterFUN = rf_filter,
                         filter_options = list(nfilter = 300),
                         family = "binomial", cores = 8, 
                         alphaSet = seq(0.7, 1, 0.05))
summary(res.rtx)

# ReliefF algorithm filter
res.rtx <- nestcv.glmnet(y = yrtx, x = data.rtx, min_1se = 0, filterFUN = relieff_filter,
                         filter_options = list(nfilter = 300),
                         family = "binomial", cores = 8, 
                         alphaSet = seq(0.7, 1, 0.05))
summary(res.rtx)
```

Leave-one-out cross-validation (LOOCV) can be performed on the outer folds.

```{r eval = FALSE}
# outer LOOCV
res.rtx <- nestcv.glmnet(y = yrtx, x = data.rtx, min_1se = 0, filterFUN = ttest_filter,
                         filter_options = list(nfilter = 300, p_cutoff = NULL),
                         outer_method = "loocv",
                         family = "binomial", cores = 8,
                         alphaSet = seq(0.7, 1, 0.05))
summary(res.rtx)
```

# Nested CV with caret

Nested CV can also be performed using the caret package framework. This enables
access to the large library of machine learning models available within caret.

Here we use caret for tuning the alpha and lambda parameters of glmnet.

```{r eval = FALSE}
# nested CV using caret
tg <- expand.grid(lambda = exp(seq(log(2e-3), log(1e0), length.out = 100)),
                  alpha = seq(0.8, 1, 0.1))
ncv <- nestcv.train(y = yrtx, x = data.rtx,
               method = "glmnet",
               savePredictions = "final",
               filterFUN = ttest_filter, filter_options = list(nfilter = 300),
               tuneGrid = tg, cores = 8)
ncv$summary

# Plot ROC on outer folds
plot(ncv$roc)

# Plot ROC on inner LO folds
inroc <- innercv_roc(ncv)
plot(inroc)
auc(inroc)

# Show example tuning plot for outer fold 1
plot(ncv$outer_result[[1]]$fit, xTrans = log)

# Extract coefficients of final fitted model
glmnet_coefs(ncv$final_fit$finalModel, s = ncv$finalTune$lambda)
```
